---
title: "Working with Database Storage Backends in `rdflib`"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval=FALSE
)
```

# Work in Progress


By default, `rdflib` stores triples in memory, which can be read in from and written out to files on disk using a variety of serializations (`rdfxml`, `json-ld`, `ntriples`, `nquads` or `turtle`).  This is analgous to reading in a `data.frame` object from a `csv`, `tsv`, or other text-delimited file, which requires the full data object to be read into memory before a user can manipulate it.  This approach runs into limitations with very large datasets, which may exceed the memory available.  Just as R packages such as `dplyr` offer the ability to perform manipulations with very large data through a connection to a database backend such as SQLite, MySQL or PostgreSQL, `rdflib` can rely on these and other databases to provide a disk-based backend. 

Unfortunately, at this time, support for these storage devices is not included in the prebuild Mac and Windows binaries of the `redland` R package.  Users wanting to take advantage of disk-based storage must instead build and install the `redland` R package from source; and in some cases, also build the `redland` C libraries from source.  This package provides a [Dockerfile]() containing a portable recipe for building the library with support for the 5 main backend storage devices: SQLite, MySQL, PostgreSQL, Virtuoso, and Berkeley DB.  This vignette documents the installation and use of these devices.  

Also see the [official documentation](http://librdf.org/docs/api/redland-storage-modules.html) for the redland C library discussing all supported storage devices.  

## Berkeley DB

The [Berkeley DB](https://en.wikipedia.org/wiki/Berkeley_DB) is a simple key-value store 

> It is the most mature and primary persistent store and suitable for large models, tested in the 2-3 million range.

Berkeley DB is the easiest of the disk-based storage options to set up.  Install both the redland libraries and the berkeley-db libraries, and then install `redland` from source.  


## Additional backends: SQLite, 

All add


`mariadb` is a drop-in replacement for MySQL (built by the original creators of MySQL after it was aquired by Oracle.)  

```{r}
rdf <- rdf(storage="mysql", host = "mariadb", 
           user="root", password="rdflib", 
           database = "mysql",
           new_db = TRUE)
```


# Checking for supported storage devices


# Testing

`rdflib` uses Circle-CI to test database backends using `docker-compose`.  `docker-compose` pulls down dedicated docker containers for `postgres` and `mariadb`, along with the `ropensci/rdflib` container, which includes a version of `redland` compiled with support for all five major backend storage systems.  At this time, testing is only configured for:

- Berkeley DB
- SQLite
- Postgres





The badge below (also on the package README) indicates that these dedicated tests are passing.

[![CircleCI](https://circleci.com/gh/ropensci/rdflib.svg?style=svg)](https://circleci.com/gh/ropensci/rdflib)
