---
title: "List queries"
author: "Carl Boettiger"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rdflib Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  
---

# Draft / work-in-progress


Often there's a JSON object/blob with some pattern I understand and want to extract without having to make explicit references to just how deep it is buried.  For instance, something like the example below:

```{r}
x <- '{
"stuff": {
   "buried": {
      "deep": [
       {
        "goodstuff": "here",
        "name": "Bob",
        "secret_power": 5
       },
       {
        "goodstuff": "here",
        "name": "Amanda",
        "secret_power": 4, 
        "more_nested_stuff": { 
            "a": 1,
            "b": 2
          }
        }
      ],
      "alsodeep": 2342423234,
      "deeper": {
        "foo": [
          {
            "goodstuff": 5,
            "name": "Barb"
          },
          {
            "goodstuff": "here",
            "name": "Scott"
          }
        ]
      }
    }
}}'
```

```{r}
blob <- jsonlite::fromJSON(x, simplifyVector=FALSE)
jq_out <- jqr::jq(x, 'recurse(.[]?) | objects | select(has("goodstuff"))')
jq_out
```

Rectangularize jq results: 


```{r}
library(magrittr)
library(jsonlite)
jq_out  %>% jqr::combine() %>% fromJSON() 
```
With `dplyr` we could then filter out the parts that we don't need (`goodstuff != "here"`), rename columns, etc.  We would have to do further work to get higher values, like the `alsodeep` code from the parent element.


```{r}

library(rdflib)
## experimental functions for rdflib package
source(system.file("examples/as_rdf.R", package="rdflib"))
blob <- as_rdf(jsonlite::fromJSON(x, simplifyVector=FALSE), vocab = "x:")
```

```{r}
q <- '
PREFIX x: <x:>
SELECT ?name, ?power ?b ?id
WHERE { 
  ?object x:goodstuff "here" .
  ?object x:name ?name .
  OPTIONAL { ?object x:secret_power      ?power } .
  OPTIONAL { ?object x:more_nested_stuff ?y .
             ?y      x:b                 ?b} .
  OPTIONAL {
             ?parent x:deep ?object .
             ?parent x:alsodeep ?id
  }
}'
rdf_query(blob, q)
```

Maybe I just want the value of the `name`s of anything found in JSON objects that also contain the key-value pair `"goodstuff": "here"`.  More generally, I'd like to just jump to that point in the "graph" and navigate from there, e.g. maybe I want to go up to the parent object and get some additional property, etc.  
