
```{r}
library(nycflights13)
library(dplyr)
library(rdflib)
```



```{r}
source(system.file("examples/as_rdf.R", package="rdflib"))
```


We need to set `rdflib` option to use disk-based rather than in-memory storage,
or it appears that `redland` throws an error (even when the machine has sufficient memory!?)
when importing the 336,776 rows of the `flights` table.  

Note: if BDB is not available (e.g. `berkeley-db` libraries were not found when `redland` was built from source),
then this will fallback on in-memory storage and this vignette will use an abridged version of the flights table.

```{r}
options(rdflib_storage = "BDB")
options(rdflib_storage = "memory")
```


## Tidyverse Style

Operations in `dplyr` on the `nyflights13` dataset are easy to write and fast to execute, (in memory or on disk):  

```{r}
df <- flights %>% 
  left_join(airlines) %>%
  left_join(planes, by="tailnum") %>% 
  select(carrier, name, manufacturer, model) %>% 
  distinct()
head(df)
```



Use a smaller dataset if we do not have a BDB backend: 

```{r}
#if(!rdf_has_bdb()){
flights <- flights %>% 
  filter(distance > 3000) # try smaller dataset
#}
```


Keys, including foreign keys, must be represented as URIs and not literal strings.  

```{r}
as_uri <- function(x, base_uri = "x:") paste0(base_uri, x)

uri_flights <- flights %>% 
  mutate(tailnum = as_uri(tailnum),
         carrier = as_uri(carrier))


```



# RDF Serialization Strategies for large data.frames

We consider a variety of strategies for actually importing `data.frames` into RDF:

- **Via `rdf_add()`**: Iterate over each row/cell with calls to `rdf_add`
- **Via JSON-LD**: Coerce the `data.frame` to JSON (via `jsonlite::toJSON(force=TRUE)`), and parse as JSON-LD
- **Via write.table()**` we `tidyr::gather()` and then hack such that we can call `write.table` on a `data.frame` to get an `nquads` text file

As we'll see, only the third solution has adequate performance here. `rdf_add()` requires an initializer call inside each `redland::addStatement`,
which takes a considerable fraction of a second.  Multiply that by the number of cells in the `data.frame` and things do not scale.

`jsonlite` can convert even the large `data.frame`s into JSON reasonably quickly. 
`jsonld::jsonld_to_rdf()` is then also acceptably fast (despite being Javascript) at converting this to `nquads`,
but unfortunately fails dramatically (i.e. `segfault`) when attempting to serialize the flights data.  (Recall we can only get into redland RDF model from JSON-LD via nquads).
Perhaps that is due to some particular data in `flights` table, but it's not obvious.  
Otherwise, this approach has lots to recommend it.  One nice feature about this approach is that it applies to almost 
any R object (e.g. any list object), though some care should be taken with names and URIs, as always. Another nice
feature is that it handles the basic data types automatically -- JSON already has types for logical, double, integer, 
and string, and these will get automatically encoded with the datatype URIs by the built-in `jsonld_to_rdf` algorithm.


The third approach is something of a poor-man's hack to the second approach.  A single call to `rdf_parse()` results in only
a single call through the redland C API to acually read in all the triples -- so unlike the `rdf_add()` approach, all the work
is being done at the C level -- the amount of R code involved doesn't at all depend on the number of triples. This is still
not nearly as fast as reading in large data.frames with `readr` or even with `read.table()`, but is probably as fast as we can get.
The trick then is to serialize the data.frame into an RDF format as quickly as possible.  We can write large `data.frame`s to text
files rather quickly with good ol `write.table()`, and after all `nquads` looks a lot like a space separated, four-column text file,
modulo a little markup to identify URIs and datatypes.  (`readr::write_delim` might be faster, but it's automatic quoting rules appear 
to be incompatible with the `nquads` use of quotations.)  We're left manually encoding the URI strings and the datatypes onto our 
`data.frame` in advance (which requires more nuiance to handle default data types, blank nodes and missing values than I've currently
implemented), but as a proof of principle here this approach is sufficiently fast, as we will now see.  


```{r}
## generic list-based conversion via JSON-LD 
rdf_planes_from_list <- as_rdf.list(planes)
```


Let's do the smaller tables first.  We declare which column is the `key` (i.e. `subject`), 
and we define a `base_uri` prefix which we use to make sure column names and subjects are treated as URIs.
With tables that have only tens of thousands of cells (triples) this is pretty fast:

```{r}
x1 <- as_rdf(airlines, "carrier", "x:")
x2 <- as_rdf(airports, "faa", "x:") ## a few funny chars, UTF8 issues?
x3 <- as_rdf(planes,  "tailnum", "x:")

x <- c(rdf(), x1,x2,x3)
```


SPARQL queries on the resulting data are also pretty fast:




```{r}
sparql <-
  'SELECT   ?model
WHERE {
 ?tailnum <x:carrier> ?carrier .
 ?tailnum <x:model>  ?model 
}'

out <- rdf_query(x1, sparql)
head(out)
```




Big table via poor-man's `nquads` 165 seconds if this is the full table:

```{r}
system.time(
    x4 <- as_rdf(uri_flights, NULL, "x:")
)
```


The json-ld approach just appears to crash, so we won't run that:

```{r}
## nope, the jsonld method appears to crash R...
#system.time(
#  x4 <- as_rdf.list(na.omit(flights))
#)
```


We can join all of these:  

```{r}
rdf <- c(rdf(), x1,x2,x3,x4)
```



Separate queries: This proves very slow on the full data! Would be much faster if we did not have to iterate over getNextResult but could parse all results as a document.  Hopefully this change is coming to `redland` R library soon!

```{r}
sparql <-
'SELECT  ?tailnum ?dep_delay ?carrier
WHERE { 
  ?flight <x:tailnum>  ?tailnum .
  ?flight <x:dep_delay>  ?dep_delay .
  ?flight <x:carrier>  ?carrier 
}'

system.time(

f1 <- rdf_query(rdf, sparql)
)
```

```{r}
sparql <-
  'SELECT  ?tailnum ?model ?manufacturer
WHERE {
?tailnum <x:manufacturer> ?manufacturer .
?tailnum <x:model> ?model
}'
f2 <- rdf_query(rdf, sparql)

tmp <- inner_join(f1,f2)
```




```{r}
s <- 
  'SELECT  ?carrier ?name ?manufacturer ?model
WHERE {
?flight <x:tailnum>  ?tailnum .
?tailnum <x:manufacturer> ?manufacturer .
?tailnum <x:model> ?model .
?flight <x:carrier>  ?carrier .
?carrier <x:name> ?name
}'

out2 <- rdf_query(rdf, s)
head(out2)
```

